---
title: "It's if_else statements all the way down..."
author: matt_gregory
comments: yes
date: '2017-02-23'
modified: `r format(Sys.time(), '%Y-%m-%d')`
layout: post
excerpt: "How to tidy up multiple if and else if statements"
published: TRUE
status: processed
tags:
- case when
categories: Rstats
output: html_document
---

```{r setup, include=FALSE}
checkpoint::setSnapshot('2016-12-22')

knitr::opts_chunk$set(
  echo = TRUE,
  dev = "svg",
  include = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  cache = TRUE
  )
```

I've been [developing](http://r-pkgs.had.co.nz/) a [package](https://github.com/mammykins) where I needed a function to take numerous different actions (different [mutations]()) depending on the values of different variables within each row of a dataframe. I started off by using a series of nested `dplyr::if_else` functions inside of a `dplyr::mutate` call. I ended up with a bit of a mess, perhaps a dozen or so `if_else` calls... that's when I got some abuse from my colleague following a Github pull request.

### Vector example

```{r 2017-02-23-barplot-fizz_if_else}
library(tidyverse)

x <- 1:50  #  a numeric vector from one to fifty
  
#  divisible by 35 with no remainder

if_else(
  x %% 35 == 0, "fizz buzz", 
  if_else(x %% 5 == 0, "fizz",
          if_else(x %% 7 == 0, "buzz",
                  "flat"
          )
  )
) %>%
  #  give the vector of character strings a nice name
  table(dnn = "fizzybuzzyness") %>%  
  #  give the frequency a nice name
  as_tibble(x = ., n = "how_many") %>%  
  #  the . means "the output piped from the previous step"
  ggplot(., aes(fizzybuzzyness, how_many)) +  
  #  strings are sorted "alphabetically"
  geom_bar(stat = "identity") +  
  govstyle::theme_gov() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))


```

There must be a better way? A quick Google got me [here](https://edwinth.github.io/blog/ifelse-wrapper/). A neat idea, but I noticed the suggestion of the new-ish `dplyr::case_when` function in the comments section of this excellent blog post. I've taken up the baton, by exploring this function in this post.  

The example from the `?case_when` is quite informative; I give a visual interpretation (see the code comments for details). However, we should explore it further and apply it to a dataframe to expand our understanding.

```{r 2017-02-23-barplot-fizz_case_when}
# ?case_when

x <- 1:50  #  a numeric vector from one to fifty

case_when(
  #  if divisible by 35 make "fizz buzz" else
  x %% 35 == 0 ~ "fizz buzz",  
  #  if divisible by 5 make "fizz", unless already "fizz buzz" else
  x %% 5 == 0 ~ "fizz",  
  #  if divisible by 7 make "buzz" unless already "fizz buzz" or "fizz"
  x %% 7 == 0 ~ "buzz",  
  #  anything else convert into flat
  TRUE ~ "flat"  
) %>%
   #  give the vector of character strings a nice name
  table(dnn = "fizzybuzzyness") %>% 
  #  give the frequency a nice name, default is n
  as_tibble(x = ., n = "how_many") %>%  
  #  the . means "the output piped from the previous step"
  ggplot(., aes(fizzybuzzyness, how_many)) +  
  geom_bar(stat = "identity") +  
  govstyle::theme_gov() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

The same outcome but which did you find the more readable? (Imagine you were quality assuring this code?)  

### Dataframes example

As pointed out in the `case_when` help examples, ordering is important where you want to go from most specific to least specific. In the example below we wanted the `Mazda RX4 Wag` to be labbelled as a `Mazda Wagon` in the newly created `brand` variable. This failed due to our ordering; to suceed we should move this before the left hand side (LHS) first argument. Notice how the right hand side of the `~` provides the replacement value. Try replacing the `"Wow"` with a numeric 50, what happens when you run the code?

```{r }
mtcars %>%
  #  convert row names to an explicit column
  tibble::rownames_to_column("thecar") %>%  
  
  
  
  #  You should start with the most specific, as ordering is important, and you
  #  can also use logical tests e.g. > <
                           
  mutate(
    brand = case_when(
      .$thecar == "Mazda RX4" | .$thecar == "Mazda RX4 Wag"  ~ "Mazda",
      .$thecar == "Mazda RX4 Wag" ~ "Mazda Wagon", 
      .$thecar == "Maserati Bora" & .$hp > 300 ~ "Wow!", 
      TRUE ~ "Not Mazda"
      )  
  ) %>%
  select(brand) %>%
  table()

```

`case_when` is still [somewhat new and experimental](https://blog.rstudio.org/2016/06/27/dplyr-0-5-0/). For now I may stick with nested `if_else` statements until this is more stable and works well within mutate despite `case_when` being a bit easier to read. If you play around with this demo code it's quite easy to break, this may be in part to some useful non-standard evaluation intrinsic to `mutate`. For example replacing the `&` with `&&` causes it to error. Try it with your own data and keep your eyes peeled for further developments! In [Hadley](https://github.com/hadley) we trust.

## Conclusion

A relatively new offering from the `tidyverse` on making nested `if_else` statements more readible.  

```{r}
devtools::session_info()
```
